---
title: "Topic 5. Visualization and Univariate Graphics"
author: "Prof. Josh Clinton"
output: beamer_presentation
fontsize: 10pt
---

# Today's Agenda

- Introduction to data visualization!

- Principles of Data Visualization

# Communicating & Visualizing Data is Essential

- Data does not exist in a vacuum -- it is always interpreted in relationship to something.

- Humans infer causality from relationships (much too quickly!).

- All data-science should be question driven!  What is the question you are answering with the visualization you are providing?

- Always think:  what is the answer that your visualization is providing?

- Goal is accuracy, not visual appeal!  Science, not Art!

- Does your visualization communicate the relationship cleanly and accurately?

# January 28, 1986: 29 Celcius
\centering
\includegraphics[height=.8\textheight]{figs/ChallIce.png}

# You are NASA (or Thokiol)...

- What is the question you are asking?

- What is the information you need to make that determination?

# Do You Launch? Pre-Launch Analysis of Damage
\centering
\includegraphics[height=.8\textheight]{figs/SlideAnalysis.png}

# Numbers are Only Meaningful in terms of Relationships
\centering
\includegraphics[height=.8\textheight]{figs/Summary.png}

# 73 seconds after launch...
\centering
\includegraphics[height=.8\textheight]{figs/Aftermath.png}

# Aftermath: Feynman
\centering
\includegraphics[height=.8\textheight]{figs/Feynman.png}

# Post Launch: Official Analysis
\centering
\includegraphics[height=.8\textheight]{figs/PostLaunchPresentation.png}

# What We Can Do Instead
```{r message=FALSE, warning = FALSE}
library(tidyverse)
library(alr4)
data(Challeng)

Betterplot <- Challeng %>%
  ggplot(aes(x=temp,y=damage)) +
  geom_point() +
  labs(title = "Relationship Between Temp. & O-Ring Damage",
       x="Termperature (Farenheit)",
       y="Measure of O-Ring Damage") + 
  scale_x_continuous(limits=c(25,85)) +
  geom_vline(xintercept=31) + 
  annotate("text",x = 31, y=10, label = "Launch Temp") +
  geom_smooth(method = "loess")
```

# What We Can Do Instead
```{r message=FALSE, warning = FALSE}
Betterplot
```

# More Info?
\centering
\includegraphics[width=.4\textwidth]{figs/Netflix.png}
\includegraphics[width=.4\textwidth]{figs/Tufte.png}

# Unfortuately...in 2003
\includegraphics[width=\textwidth]{figs/ColumbiaExplosion.png}

# Unfortuately...in 2003
\includegraphics[width=\textwidth]{figs/Columbia2003.png}


# GOAL: Your visuals must tell an accurate story

- Humans over-interpret relationships as causal.  You must be informative without being misleading.

- Tables and graphs are essential.

- Visualizations must be stand-alone as much as possible.

NOTE: Rule of thumb: show it to someone without explanation. If they are confused, re-do!

# Dimensions of Visualization

Essential elements of every good visualization:

- Well-labeled axes (no variable names!).

- Title describing the data being plotted.

- Meaningful scale for x and y axes.

- Appropriate visualization for the type of data being characterized.

- ADVANCED: Uncertainty (if relevant)

# Dimensions of Visualization

You have several "dimensions" to use when presenting information

- Horizontal dimension (x-axis) 

- Vertical dimension (y-axis) 

- Shape of data points

- Color

- Size of data points

# Don't be confusing

- Map each variable to at most one dimension.  

- Be intuitive -- don't assign small numbers large dots, etc.

- Don't include extraneous information/visual elements.

# Best Graph Ever?  Charles Joseph Minard

\includegraphics[height=.95\textheight]{figs/Minard}

# Or maybe John Snow?  (Not Jon Snow!)

\includegraphics[height=.95\textheight]{figs/Snow}

# Your graphs should make sense!

- The visualization should clearly answer the question.

\centering
\includegraphics<1>[height=.9\textheight]{figs/super-bowl}
\includegraphics<2>[height=.9\textheight]{figs/salaries}

# Your graphs should make sense!
- The visualization itself should not be the focus of the visualization.  The focus should be on the relationship being visualized.

\includegraphics[height=\textheight]{figs/infographic}

# Your graphs should make sense!
- The visualization should be complete (present all the information) but not include extraneous/distracting information/visual-elements.

\includegraphics[height=\textheight]{figs/NBCMap}

# Your graphs should be accurate and not mislead

- Numbers (and patterns in numbers) should match the visualization

\centering
\includegraphics<1>[height=.9\textheight]{figs/sun-pie}
\includegraphics<2>[height=.9\textheight]{figs/temps}
\includegraphics<3>[height=.9\textheight]{figs/budget}
\includegraphics<4>[height=.9\textheight]{figs/drugs}
\includegraphics<5>[height=.9\textheight]{figs/NovaScotiaVotes}

# Your graphs should be accurate and not mislead
- Avoid piecharts, but if you use them make sure they sum to 100.

\includegraphics<1>[height=.9\textheight]{figs/Fox100}
\includegraphics<2>[height=.95\textheight]{figs/bitcoin}

# Your graphs should be accurate and not mislead
- Variation in a variable should vary in one-visual dimension, not two! (Piecharts bad)

\centering
\includegraphics<1>[width=\textwidth]{figs/battery}
\includegraphics<2>[height=.8\textheight]{figs/cake}

# But, don't put too much info on a graph!

- Think about the central point you are trying to convey!  

- A visualization with too much data (and not enough variation in that data?) is meaningless...

\centering
\includegraphics<1>[height=.85\textheight]{figs/too-much-1}

- As are graphs with too little variation

# Don't use a graph if you don't need a graph!
\centering
\includegraphics<1>[width=\textwidth]{figs/mueller}
\includegraphics<2>[height=.9\textheight]{figs/lgbt}
\includegraphics<3>[width=\textwidth]{figs/drinking-canada}

# Barplots are good, but not 3d barplots!

- Does the visualization reveal all of the variation of interest?

- Depth is a dimension that can show variation, but it can be hard to interpret!

\includegraphics[width=\textwidth]{figs/3dbar}

# Don't use deceptive y-axes...

- We don't always need to include 0, unless not doing so is confusing.

\includegraphics[height=.9\textheight]{figs/heights}

# Don't use deceptive y-axes...

- Axes can make small changes look big and big changes look small.

\includegraphics[width=\textwidth]{figs/border}

# Don't use deceptive y-axes...

- The range can also change how you interpret the data in relationship to suggested relationships.

\includegraphics[width=\textwidth]{figs/welfare}

# ...But please do use axes!
\includegraphics<+>[width=\textwidth]{figs/obamacare}
\includegraphics<+>[width=\textwidth]{figs/SoccerXaxis}

# But not multiple y-axes!
\centering
\includegraphics<+>[height=.9\textheight]{figs/ads}
\includegraphics<+>[height=.8\textheight]{figs/vinyl1}
\includegraphics<+>[height=.8\textheight]{figs/vinyl2}
\includegraphics<+>[width=\textwidth]{figs/bubbles}
\includegraphics<+>[width=\textwidth]{figs/TechTesla}

# Only use colors if meaningful
\centering
\includegraphics<+>[height=.9\textheight]{figs/obama}
\includegraphics<+>[height=.9\textheight]{figs/operating-systems}

# Uncertainty?  
- How confident are we about our visualized predictions?
\centering
\includegraphics<+>[height=.9\textheight]{figs/Bills}

# Things to think about

- What is the goal of the graph -- what information do you want to convey and for what purpose?

- What is your eye drawn to?  Is that relevant?

- What does your brain tell you about what you conclude from that?  Is that the correct conclusion?

- Visualization is reducing complexity (hopefully), but is the reduction informative or confusing?

- Graphs overwhelm text and tables so think about them!

#   Is this a good visualization of the pandemic in the US?

\includegraphics[height=.9\textheight]{figs/CovidMap}

https://coronavirus.jhu.edu/us-map

# How about these?

\includegraphics<+>[height=.9\textheight]{figs/RiskOverTime}
\includegraphics<+>[height=.9\textheight]{figs/RiskFeb7}
\includegraphics<+>[height=.9\textheight]{figs/CovidVU}

# Visualization is very powerful

- Humans infer causality very quickly so make sure that the relationships are not misguided.

- Make sure people get what you intend from the visualization.
